Staged Abstract Interpreters

### 2019-09-08 paper/10_intro.tex
Proofreading intro

Abstract interpretation as a lattice-based approach to sound static analyses was
proposed by \citet{DBLP:conf/popl/CousotC77, CousotCousot79-1}. Based on the notion of Galois
connections, the analyzer can soundly approximate the
concrete program behavior at runtime by computing fixed points on an abstract
domain. Despite the tremendous theoretical development of abstract
interpretation over the years, constructing artifacts and analyzers that perform
sound abstract interpretation for modern and expressive languages was considered
complicated for a long time.

Recent progress on methodologies such as Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an
operational approach to constructing abstract interpreters.  By deriving
semantics artifacts for abstract interpretation from their concrete
counterparts (for example, abstract machines), soundness can be easily
established by examining the transformation of semantic artifacts.
This systematic abstraction approach can be tailored to different
language features (such as state, first-class control, exceptions, and
concurrency) and sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

@@

Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Kiselyov:2015:FMM:2804302.2804319, Githubsemantic}, etc. These pure
approaches provide certain benefits, for example, the abstract interpretation
artifacts can be built in a compositional and modular fashion, and referential
transparency allows programmers to confidently reason about the correctness of
their implementations.  The soundness of an analysis can also be proven with less
effort, either by mechanized proofs \cite{Darais:2016:CGC:2951913.2951934} or
paper-based proofs \cite{Keidel:2018:CSP:3243631.3236767}.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs
interpretive overhead, such as pattern matching on the AST nodes and recursive
calls on subexpressions. If it is a single-time traversal, such overhead can be
negligible. But usually, abstract interpreters iteratively analyze and traverse
the AST multiple times to reach fixed-points, therefore the accumulated
overhead is even magnified. 
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).  
Second, it is common that abstract interpreters written in pure languages
extensively use functional encodings of side-effects to implement the semantics
of abstract interpretation.  For example, the list monad can be used to model
nondeterminism from collecting interpreters.
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed,
\citet{futamura1971partial} observed a close
connection between interpreters and compilers through a hierarchy of
specializations, since then known as Futamura Projections. The first Futamura
projection describes that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. 
In recent years, the idea of Futamura projections has attracted more attentions
and has been successfully applied to many real-world problems, including
query compilers \cite{DBLP:conf/sigmod/TahboubER18}, PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, etc.
The challenge here: can we apply Futamura projections to a general abstract interpreter
to make it run faster? The solution is not obvious, as we need to specialize
a nonstandard semantics that extensively uses nondeterminism, operates on
abstract domains, and computes fixed-points iteratively over caches.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework that eliminates performance penalties for monadic abstract
interpreters, while keeping the implementation as close to their conceptual
model as possible.
In short, we borrow ideas and techniques from multi-stage programming and
embedded DSLs, and apply them to abstract interpreters:  1) To
remove the overhead from interpretation and effect layers, we specialize the
abstract interpreter via staging and then generate efficient low-level code
that performs the actual analysis. 2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-times and different semantics, which allows the staged
abstract interpreter to be derived from its unstaged counterpart.  As a result,
the derived staged abstract interpreter has no intrusive changes to the
underlying abstract semantics, thereby preserving soundness.  In this
sense, our approach allows for no regret for both performance and engineering
effort. We elaborate these two main ideas in detail, as follows.

@@

\paragraph{Futamura Projection of Abstract Interpreters}

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent executable. For
instance, assume that @eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization, we can
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply Futamura projections on abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract
interpreter as the unstaged baseline.  Similar to the two-level semantics
\cite{NIELSON1989117}, monads provide a good abstraction to hide the detail of
abstract interpretation semantics. However, monads do not have a clear
stage distinction.
We further introduce binding-times and make the monads binding-time polymorphic.
Our proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection
of an abstract interpreter for a stateful higher-order
language. By deriving staged monads that can be used to perform code generation, we
obtain a staged abstract interpreter that can be used for specialization. Through
staging, the generated code is specialized to the input program, and all the
monadic operations are inlined and compiled down to low-level Scala code.

>>>>

Abstract interpretation as a lattice-based approach to sound static analyses was
proposed by \citet{DBLP:conf/popl/CousotC77, CousotCousot79-1}. Based on the
notion of Galois connections, an analyzer can soundly approximate concrete
program behavior at runtime by computing fixed points on an abstract domain.
Despite the tremendous theoretical development of abstract interpretation over
the years, constructing artifacts and analyzers that perform sound abstract
interpretation for modern and expressive languages was considered complicated
for a long time.

Recent progress on methodologies such as Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an operational
approach to constructing abstract interpreters.  By deriving semantics artifacts
for abstract interpretation from their concrete counterparts (for example,
abstract machines), soundness can be easily established by examining the
transformation of semantic artifacts. This systematic approach to abstraction
can be tailored to different language features (such as state, first-class
control, exceptions, concurrency, etc.) and sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
Darais:2015:GTM:2814270.2814308}. It has also been applied to various small-step
abstract machines \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
Sergey:2013:MAI:2491956.2491979} and big-step definitional interpreters
\cite{Wei:2018:RAA:3243631.3236800, DBLP:journals/pacmpl/DaraisLNH17,
Keidel:2018:CSP:3243631.3236767}.

@@

Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows \cite{Keidel:2018:CSP:3243631.3236767},
extensible effects \cite{Kiselyov:2015:FMM:2804302.2804319, Githubsemantic},
etc. These pure approaches provide certain benefits, including the fact that the
abstract interpretation artifacts can be built in a compositional and modular
fashion. Furthermore, referential transparency allows programmers to confidently
reason about the correctness of their implementations.  The soundness of an
analysis can also be proven with less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of abstract interpretation.  For example, the list monad can be used to model
nondeterminism from collecting interpreters. Although such pure approaches have
their own merits, they are significantly slower when compared with imperative,
stateful implementations.  In order to reconcile these approaches, one must use
a programming technique which allows for high-level abstractions during
implementation, but bypasses the incurred overhead at runtime.

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura Projections. The first Futamura projection states that specializing an
interpreter to an input program removes interpretative overhead and yields an
equivalent executable that runs faster. In recent years, the idea of Futamura
projections has attracted more attention, and has been successfully applied to
many real-world problems, including query compilers
\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, etc. The challenge here: can we
apply Futamura projections to a general abstract interpreter to make it run
faster? The solution is not obvious, as we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-points iteratively over caches.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework that eliminates performance penalties for monadic abstract
interpreters, while keeping the implementation as close to their conceptual
model as possible. In short, we borrow ideas and techniques from multi-stage
programming and embedded DSLs, and apply them to abstract interpreters:  1) To
remove the overhead from interpretation and effect layers, we specialize the
abstract interpreter via staging, and then generate efficient low-level code
that performs the actual analysis. 2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-times and different semantics, which allows the staged
abstract interpreter to be derived from its unstaged counterpart.  As a result,
the derived staged abstract interpreter has no intrusive changes to the
underlying abstract semantics, thereby preserving soundness.  In this sense, our
approach allows for a complete absence of regret in terms of both performance
and engineering effort. We elaborate these two main ideas in detail, as follows.

@@

\paragraph{Futamura Projections Meet Abstract Interpreters}

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent executable. For
instance, assume that @eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization we can
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The first proposed approach to realizing Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes between compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program
over a two-level syntax \cite{Nielson:1992:TFL:130665} by manually annotating
the binding-times of variables. The MSP system will then check whether these
annotations are consistent, and specialize the program using that information.
The staging annotations can be either syntactic (e.g., quote and quasi-quote in
MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract interpreter
as the unstaged baseline.  Similar to two-level semantics \cite{NIELSON1989117},
monads provide a good abstraction to hide the detail of abstract interpretation
semantics. However, monads do not have a clear stage distinction. We further
introduce binding-times and make the monads binding-time polymorphic. Our
proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection of
an abstract interpreter for a stateful higher-order language. By deriving staged
monads that can be used to perform code generation, we obtain a staged abstract
interpreter that can be used for specialization. Through staging, the generated
code is specialized to the input program, and all the monadic operations are
inlined and compiled down to low-level Scala code.
